---
title: "spicyR: calibrated tests for changes in cell-type co-localisation"
date: "`r BiocStyle::doc_date()`"
author:
- name: Ellis Patrick
  affiliation:
  - School of Mathematics and Statistics, University of Sydney, Australia
  - Westmead Institute for Medical Research, University of Sydney, Australia
- name: Sadiq Dohadwalla
  affiliation: School of Mathematics and Statistics, University of Sydney, Australia
- name: Elijah Willie
  affiliation: School of Mathematics and Statistics, University of Sydney, Australia
- name: Nicolas Canete
  affiliation: Westmead Institute for Medical Research, University of Sydney, Australia
package: "`r BiocStyle::pkg_ver('spicyR')`"
abstract: >
  spicyR tests whether the spatial co-localisation of two cell types changes between groups of patients, or is
  associated with patient outcome. Its default method, spicyR Cell, counts the cells of one type around each
  cell of another, compares the count with its exact expectation under random labelling of the observed
  cells, and tests the difference between groups with patients, not images or cells, as the units of
  inference. This vignette analyses about 400,000 cells from several hundred breast cancer patients.
vignette: >
  %\VignetteIndexEntry{spicyR: calibrated tests for changes in cell-type co-localisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    toc_float: true
    toc_depth: 2
bibliography: REFERENCES.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 4.5, fig.align = "center")
```

# Overview

Spatial omics technologies (imaging mass cytometry, CODEX, MIBI, Xenium, CosMx and others) record the type and
location of every cell in a tissue section. A natural question is whether two cell types sit closer together,
or further apart, in one group of patients than in another: do T cells crowd around tumour cells more in
responders than in non-responders?

`spicy()` answers this question for every pair of cell types at once. For a pair written **from → to**, it
asks how many **to** cells lie within a radius *r* of each **from** cell, and whether that number changes
between conditions.

Three features make the answers trustworthy:

1. **The comparison is with random labelling of the observed cells.** For each image, the count of *to* cells
   around *from* cells is compared with its exact expectation if the cell-type labels were shuffled among the
   cells that are there. Tissue shape, holes, folds and cell density are conditioned on, not modelled. The effect
   is the **excess**: the number of extra *to* cells per *from* cell, beyond chance.
2. **Patients are the units.** Images are combined within patients and patients within conditions. The
   difference between conditions is tested with a small-sample cluster-robust (CR2) variance on Satterthwaite
   degrees of freedom. Images and cells are never treated as independent replicates, so a patient contributing
   ten images does not count ten times.
3. **Abundance is kept apart from attraction.** The excess of a pair can change because the counted type
   becomes more or less common, even if cells are arranged no differently. Every result therefore also reports
   the difference at equal availability of the counted type.

The original spicyR test, which compares per-image L-function summaries with linear mixed models
[@canete2022spicyr], is still available with `method = "image"` and has its own vignette.

# Installation

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install("spicyR")
```

```{r packages}
library(spicyR)
library(SpatialDatasets)
library(survival)
library(ggplot2)
```

# Data: METABRIC imaging mass cytometry

We use the imaging mass cytometry data of @ali2020imaging: tumour cores from the METABRIC breast cancer cohort,
with cell types assigned by the authors, one core per patient. It is available from the `SpatialDatasets`
package. A data set with several images per patient follows in the section on patients as units.

```{r data}
spe <- spe_Ali_2020()
cells <- as.data.frame(SummarizedExperiment::colData(spe))
cells$imageID <- cells$file_id
cells$cellType <- cells$description
cells$x <- cells$Location_Center_X
cells$y <- cells$Location_Center_Y
cells <- cells[cells$ER.Status %in% c("neg", "pos"), ]
cells$ER <- factor(cells$ER.Status, levels = c("neg", "pos"))
```

```{r data-summary}
c(cells = nrow(cells), images = length(unique(cells$imageID)),
  patients = length(unique(cells$metabricId)), cell_types = length(unique(cells$cellType)))
table(unique(cells[, c("metabricId", "ER")])$ER)   # patients per group
```

`spicy()` also accepts `SingleCellExperiment` and `SpatialExperiment` objects directly; the column names are
set with the `imageID`, `cellType` and `spatialCoords` arguments.

```{r plot-image, fig.height = 5}
one <- cells[cells$imageID == cells$imageID[1], ]
ggplot(one, aes(x, y, colour = cellType)) +
  geom_point(size = 0.6) +
  coord_equal() +
  labs(title = one$imageID[1], colour = NULL) +
  theme_void() +
  theme(legend.text = element_text(size = 7))
```

# Testing every pair of cell types

The only choices are the column of the condition, the column of the patient (`subject`) and the radius.
Coordinates here are in micrometres, and 25 µm is about two to three cell diameters.

```{r spicy}
t0 <- Sys.time()
res <- spicy(cells, condition = "ER", subject = "metabricId", r = 25)
Sys.time() - t0
res
```

All `r nrow(res$cellResults)` ordered pairs of `r length(unique(cells$cellType))` cell types, from
`r format(nrow(cells), big.mark = ",")` cells, take a few seconds. `topPairs()` lists the most significant pairs. `coefficient` is the change in the
excess between ER-negative (the reference level) and ER-positive patients, in extra *to* cells per *from*
cell, and `intercept` is the excess in the reference group.

```{r top-pairs}
topPairs(res, n = 10)
```

The full table, `res$cellResults`, has one row per pair:

- `excess_ref`, `excess_comp`: the excess in each condition;
- `excess_difference`, `se`, `df`: their difference, its standard error and degrees of freedom;
- `p_value`, `p_adj`: the p-value, and the Benjamini–Hochberg adjusted p-value over all pairs;
- `tau2`: the variance of the patient frailty;
- `adjusted_*`: the same quantities at equal availability (next section).

```{r results-table}
head(res$cellResults[order(res$cellResults$p_value), c("from", "to", "excess_ref", "excess_comp",
  "excess_difference", "se", "df", "p_value", "p_adj")])
```

`signifPlot()` shows every pair at once.

```{r signif-plot, fig.width = 9, fig.height = 8}
signifPlot(res, breaks = c(-3, 3, 0.5))
```

# Looking at one pair

`spicyBoxPlot()` shows the excess in each image, so you can see what drives a result.

```{r box-plot}
top <- topPairs(res, n = 1)
spicyBoxPlot(res, from = top$from, to = top$to)
```

`plotImage()` shows the two cell types of a pair in one image, over the density of all cells.

```{r plot-pair, fig.height = 5}
SummarizedExperiment::colData(spe)$imageID <- spe$file_id
SummarizedExperiment::colData(spe)$cellType <- spe$description
plotImage(spe, imageToPlot = cells$imageID[cells$ER == "pos"][1], from = top$from, to = top$to)
```

`bind()` returns the same per-image values as a table, ready for any other analysis.

```{r bind}
head(bind(res)[, 1:5])
```

# Abundance or attraction?

The top pairs above involve hormone-receptor-positive tumour cells (`HR+ CK7-`), which are much more common in
ER-positive tumours. When a type becomes more common, more of its cells are found around any other cell, even
if the arrangement is unchanged. The `adjusted_*` columns answer a sharper question: does the excess differ
between the groups **at equal availability of the counted type**? The adjustment adds the log share of the
counted type in each image to the model, with a slope shared by the two groups.

```{r availability}
both <- res$cellResults[, c("from", "to", "excess_difference", "p_adj", "adjusted_difference", "adjusted_p_adj")]
both <- both[order(both$p_adj), ]
head(both, 10)
c(raw = sum(both$p_adj < 0.05), adjusted = sum(both$adjusted_p_adj < 0.05, na.rm = TRUE))
```

```{r availability-plot}
ggplot(both, aes(-log10(p_adj), -log10(adjusted_p_adj))) +
  geom_abline(linetype = 2, colour = "grey50") +
  geom_point(alpha = 0.6) +
  labs(x = "-log10 BH p (raw excess)", y = "-log10 BH p (at equal availability)") +
  theme_classic()
```

Pairs close to the diagonal are robust to the abundance of the counted type. Pairs that fall away from it
changed mainly because the counted type became more or less common. Both are reported because both can be of
interest: in a disease where composition change is the biology, the raw excess may be the question.

# Several images per patient

When a patient contributes several images, the images are not independent replicates: they share the patient's
biology. Give the patient column as `subject` and the patients become the units. To see why this matters, take
the colorectal cancer CODEX data of @schurch2020coordinated, with about four tissue microarray spots from each of
35 patients in two groups, and compare the analysis with patients as units against one that treats every spot
as a patient. A `SpatialExperiment` can be passed directly.

```{r schurch}
schurch <- spe_Schurch_2020()
# the study's two patient groups: Crohn's-like reaction (CLR, group 1) and diffuse inflammatory infiltration (DII, group 2)
schurch$group <- factor(c("CLR", "DII")[as.integer(as.character(schurch$group))], levels = c("CLR", "DII"))
table(table(unique(as.data.frame(SummarizedExperiment::colData(schurch))[, c("imageID", "patient")])$patient))
byPatient <- spicy(schurch, condition = "group", subject = "patient", r = 50)
bySpot <- spicy(schurch, condition = "group", r = 50)       # every spot treated as a patient (wrong)
byPatient
bySpot
```

```{r schurch-df}
d <- merge(byPatient$cellResults[, c("from", "to", "df", "p_value")],
           bySpot$cellResults[, c("from", "to", "df", "p_value")], by = c("from", "to"), suffixes = c("_patients", "_spots"))
summary(d[, c("df_patients", "df_spots")])
```

With spots as units the test has roughly four times the degrees of freedom it is entitled to, and its p-values
are too small. With patients as units the degrees of freedom are bounded by the number of patients, as they
should be.

# Covariates

Covariates measured per image or per patient are added to the model with `covariates`. Here the difference
between ER groups is adjusted for age at diagnosis and tumour grade.

```{r covariates}
cells$Grade <- factor(cells$Grade)
resCov <- spicy(cells, condition = "ER", subject = "metabricId", r = 25, covariates = c("Age.At.Diagnosis", "Grade"),
                from = c("T cells", "Macrophages Vim+ Slug-"), to = c("HR+ CK7-", "HR- CK7+", "T cells"))
resCov$cellResults[, c("from", "to", "excess_difference", "p_value", "covariate_difference", "covariate_p_value")]
```

# Several radii

The scale of an interaction is rarely known in advance. Give several radii and `spicy()` combines them with a
max-T test that accounts for the strong correlation between nested radii (`combine = "cauchy"` is an
alternative). The reported effect is the one at the radius with the strongest evidence.

```{r radii}
resR <- spicy(cells, condition = "ER", subject = "metabricId", r = c(10, 25, 50),
              from = c("T cells", "B cells"), to = c("T cells", "B cells", "HR+ CK7-"))
resR$cellResults[, c("from", "to", "r", "excess_difference", "p_value", "p_value_best_radius")]
```

```{r radii-plot}
ggplot(resR$radiusResults, aes(r, excess_difference, colour = paste(from, "->", to))) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
  geom_line() + geom_point() +
  labs(x = "radius (µm)", y = "difference in excess (ER+ minus ER-)", colour = "pair") +
  theme_classic()
```

# Survival

A `Surv` column as the condition tests whether co-localisation is associated with outcome. The test regresses
each patient's excess on the martingale residual of a Cox model without spatial terms (a score test that does
not need the Cox model to hold). It also reports the hazard ratio per standard deviation of the
patient-level excess, estimated after shrinkage of each patient's excess towards the mean.

```{r survival}
cells$RFS <- Surv(cells$timeRFS, cells$eventRFS)
resS <- spicy(cells, condition = "RFS", subject = "metabricId", r = 25, covariates = "Age.At.Diagnosis")
head(resS$cellResults[order(resS$cellResults$p_value),
  c("from", "to", "score_coefficient", "p_value", "p_adj", "hazard_ratio_sd", "hr_p_value")])
```

# Options

- `k = 10` counts the 10 nearest neighbours instead of a radius.
- `variance = "hartung_knapp"` is for studies with very few patients per group. It uses the model-based variance,
  floored at CR2, on *m* − 2 degrees of freedom.
- `labelClustering = FALSE` turns off the inflation of the within-image variance for target cells that cluster
  among themselves (on by default).
- `ref` chooses the reference condition.
- `cores` sets the number of threads.

```{r knn}
resK <- spicy(cells, condition = "ER", subject = "metabricId", k = 10, from = "T cells", to = c("HR+ CK7-", "B cells"))
resK$cellResults[, c("from", "to", "excess_difference", "p_value")]
```

# The original image-level test

`method = "image"` runs the original spicyR test: a per-image summary of the L function of each pair, compared
between conditions with a weighted linear mixed model with a random intercept for each patient
[@canete2022spicyr]. It remains useful with alternative per-image statistics (`alternateResult`), for example
from Kontextual. See the vignette "The original image-level test in spicyR".

```{r image-method}
resImg <- spicy(cells, condition = "ER", subject = "metabricId", method = "image",
                from = "T cells", to = c("HR+ CK7-", "B cells"))
topPairs(resImg)
```

# Python

The Python package [`spicyr`](https://github.com/SydneyBioX/spicyr-py) runs the same analysis from SpatialData,
AnnData or pandas objects, on the same C++ core. The two packages give the same results to numerical precision.

```{python, eval = FALSE}
import spicyr
res = spicyr.spicy(adata, condition="ER", subject="metabricId", r=25)
res.top_pairs(10)
```

# Session information

```{r session}
sessionInfo()
```

# References
